A Computational Account of Default-Mode Function\\ by Control Theory and Reinforcement Learning

### 2016-09-29 paper/paper.tex
a note on model-based RL + more

\section{Reinforcement-Learning as a computational model for DMN function}
The main idea is to show that the DMN has biological substrates (gradient
computation, backprop, experience replay, etc.) for all the
major components of an efficient RL algorithm, and all these components
integrate in a plausible way.
\subsection{General Reinforcment-Learning (RL) theory}
In a general reinforcement-learning (RL) framework, an \textit{agent}
interacts with an unknown environment in trial-and-error fashion by taking
\textit{actions / decisions} $a_t$ which trigger changes in the
\textit{state} of the environment
$s_t \rightarrow s_{t+1}$, accompanied with a real-valued \textit{reward}
$r_t$ (or regret!, since it can be negative). This reward should be thought
of as an instantaneous satisfaction accompanying the execution of
an action. Here the subscript $t$ denotes time, and can be discrete or
continuous.
%% The unknown environment $\mathcal E$ is generally taken as stochastic, but having some kind of
%% memory.
A popular abstract model for the scenario are so-called \textit{Markov Decision Processes (MDPs)}.
This is simply a quintuplet $(\mathcal S, \mathcal A, r, \rho, \mu)$ where
\begin{itemize}
\item $\mathcal S$ is the set of states, e.g $\mathcal S = \{\text{happy}, \text{sad}\}$.
\item $\mathcal A$ is the set of actions, e.g $\mathcal A = \{\text{read}, \text{run},
  \text{laugh}, \text{sympathize}, \text{empathize}\}.$
\item $r : \mathcal S \times \mathcal A \rightarrow \mathbb R$, so that $r(s, a)$ is the instant reward for taking action $a$ in state $s$.
\item $\mu: \mathcal S \rightarrow [0, 1]$ is the prior probability on the states so that
  $\mu(s)$ is the probability that the environment starts-off in state $s$.
  \item $\rho : \mathcal S \times \mathcal A \times \mathcal S \rightarrow [0, 1]$, so that
    $\rho(s,a,s') = \mathbb P(s'|s, a)$ is the probability of moving to state $s'$ if action $a$ is taken from state $s$. In addition, one requires that such transitions be Markovian.
\end{itemize}

@@

The behavior of the agent is governed by some kind of \textit{policy}, which maps states to
sampling probabilities over candidate actions to perform from thin state. The  goal of RL is then
adapt this policy so as to  maximize the cumulative rewards.

@@

\textit{Q-learning} (Watkins \& Dayan, 1992), a commonly used off-policy \textit{critic-only} algorithm, uses the
\textit{greedy policy}
\begin{equation}
  \pi(s) = \pi_{\tilde{Q}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a),
  \label{eq:qlearning}
\end{equation}
where $\tilde{Q}(s, a)$ is the current estimate of the state-action value table.
The hallmark of this trick wherein one only considers deterministic policies is the so-called
\textit{Bellman's equation} [Sutton and Barto 1998]
\begin{equation}
  Q^\pi(s_t, a_t) = \mathbb E_{s_{t+1} \sim \rho(.|s_t)} [r(s_t,a_t) + \gamma Q^\pi(s_{t+1}, \pi(s_{t+1}))]
  \label{eq:bellman}
\end{equation}
which is a fixed-point equation for the deterministic policy $\pi$.
Notice that in the expectation in \eqref{eq:bellman}, the sampling is done over only things which
depend on the environment, and so can be learned off-policy by observing state-transitions
triggered by another behavioral policy, which can be stochastic.

\subsection{Q-learning with value approximation and replay}
As already mentioned in the previous section, Q-learning optimizes over the class of
deterministic policies of the form \eqref{eq:qlearning}.
in state $s \in \mathcal S$, the agent takes an action $a \in \mathcal A$
% (for example, $a$ = ``Read'', or ``Run'', or ``Laugh'', or ``Sympathize'', or ``Empathize'', etc.)
by sampling from its current policy matrix, and collects a reward $r$,
% (measured on their internal currency of worldly values ?),
and the environment transitions to a new $s' \in \mathcal S$. At the end of
this cycle, a new \textit{experience} $e = (s,a,r,s')$ is produced; this represents an exemplar behavior of the agent and is recorded in replay memory:
$\mathcal D \leftarrow \mathcal D \cup \{e\}$.
Now, replay consists in sampling (uniform or importance-weighted ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal U(D)$ from the replay memory $\mathcal D$ and trying to
approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by Bellman's equation \eqref{eq:bellman}, namely
$r + \gamma \max_{a'} Q(s', a'|\theta^Q)$, with the output of a parametrized regression model $(s,a,r,s') \mapsto {Q}(s, a|\theta^Q)$, where $Q(.,.|\theta^Q)$ is an approximator for the $Q$-value operator,  parameterized by $\theta^Q$. Computing an optimal policy then corresponds to finding the parameters $\theta^Q$ which minimize the following loss function
\begin{equation}
  \mathcal L(\theta^Q)
  % = \mathbb E_{(s, a, r, s') \sim U(\mathcal D)}[(Q(s, a|\theta^Q) - Q(s',a'|\theta^Q))^2]
  = \mathbb E_{(s, a, r, s') \sim U(\mathcal D)}[(Q(s, a|\theta^Q) - y)^2],
  \label{eq:oracle}
  \end{equation}
where
$ y = r + \gamma \max_{a' \in \mathcal A} Q(s', a'|\theta^Q) = r + \gamma Q(s', \pi(s')|\theta^Q)$ is the prediction target.
% While $y_t$ is also dependent on $\theta^Q$, this is typically ignored.

>>>>

\section{Reinforcement-Learning as a computational model for DMN function}
The main idea is to show that the DMN has biological substrates (gradient
computation, backprop, experience replay, etc.) for all the
major components of an efficient RL algorithm, and all these components
integrate in a plausible way.
\subsection{Markov Decision Processes}
In a general reinforcement-learning (RL) framework, an \textit{agent}
interacts with an  environment in a trial-by-error fashion by taking
actions / decisions $a$, which trigger changes in the
\textit{state} of the environment
$s \rightarrow s'$, accompanied with a real-valued \textit{reward}
$r = r(s, a)$ (or regret!, since it can be negative) collected by the
agent. This reward should be thought
of as an instantaneous satisfaction accompanying the execution of
an action. Here the subscript $t$ denotes time, and can be discrete or
continuous.
%% The environment $\mathcal E$ is generally taken as stochastic, but having some kind of
%% memory.
A popular abstract model for the scenario are so-called \textit{Markov Decision Processes (MDPs)}.
This is simply a quintuplet $(\mathcal S, \mathcal A, r, \rho, \mu)$ where
\begin{itemize}
\item $\mathcal S$ is the set of states, e.g $\mathcal S = \{\text{happy}, \text{sad}\}$.
\item $\mathcal A$ is the set of actions, e.g $\mathcal A = \{\text{read}, \text{run},
  \text{laugh}, \text{sympathize}, \text{empathize}\}.$
\item $r : \mathcal S \times \mathcal A \rightarrow \mathbb R$, so that $r(s, a)$ is the instant reward for taking action $a$ in state $s$.
\item $\mu: \mathcal S \rightarrow [0, 1]$ is the prior probability on the states so that
  $\mu(s)$ is the probability that the environment starts-off in state $s$.
  \item $\rho : \mathcal S \times \mathcal A \times \mathcal S \rightarrow [0, 1],\; (s,a,s') \mapsto \rho(s'|s,a)$, the probability of moving to state $s'$ if action $a$ is taken from state $s$. In addition, one requires that such transitions be Markovian.
\end{itemize}

@@

\paragraph{A note on model-based RL.}
Knowledge the agent has about the environment can be encoded in the
functions $r$, $\mu$, $\rho$, and / or $\pi$, as is done in
\textit{model-based} RL. For example if a rat in a maze knows, for sure,
that staying still will produce no change in the environment, and in particular will not eventually lead to finding the food, they can assume:
\begin{itemize}
\item $r(s, \text{``stand still''}) = 0$ if $s$ does not correspond to a cell / chamber containing food.
\item $\rho(s'|s,\text{``stand still''}) = 1$ if $s'=s$ and $0$ otherwise.
\item etc.
\end{itemize}

\subsection{Value functions and policies}
The behavior of the agent is governed by some kind of \textit{policy}, which maps states to
sampling probabilities over candidate actions to perform from thin state. The  goal of RL is then
adapt this policy so as to  maximize the cumulative rewards.

@@

\textit{Q-learning} (Watkins \& Dayan, 1992), a commonly used off-policy algorithm, uses the
\textit{greedy policy}
\begin{equation}
  \pi(s) = \argmax_{a \in \mathcal A}{Q}(s, a).
  \label{eq:qlearning}
\end{equation}
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
The celebrated \textit{Bellman's equation} [Sutton and Barto 1998] then takes the simple form
\begin{equation}
  Q^\pi(s_t, a_t) = \mathbb E_{s_{t+1} \sim \rho(.|s_t)} [r(s_t,a_t) + \gamma Q^\pi(s_{t+1}, \pi(s_{t+1}))]
  \label{eq:bellman}
\end{equation}
which is a fixed-point equation for the deterministic policy $\pi$.
Notice that in the expectation in \eqref{eq:bellman}, the sampling is done over only things which
depend on the environment, and so can be learned off-policy by observing state-transitions
triggered by another behavioral policy, which can be stochastic.

\subsection{Efficient RL with value approximation and replay}
As already mentioned in the previous section, Q-learning optimizes over the class of
deterministic policies of the form \eqref{eq:qlearning}.
in state $s \in \mathcal S$, the agent takes an action $a \in \mathcal A$
% (for example, $a$ = ``Read'', or ``Run'', or ``Laugh'', or ``Sympathize'', or ``Empathize'', etc.)
by sampling from its current policy matrix, and collects a reward $r$,
% (measured on their internal currency of worldly values ?),
and the environment transitions to a new $s' \in \mathcal S$. At the end of
this step, a new \textit{experience} $e = (s,a,r,s')$ is produced; this represents an exemplar behavior of the agent and is recorded in replay memory:
$\mathcal D \leftarrow \mathcal D \cup \{e\}$.
Now, replay consists in sampling (uniform or importance-weighted\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal U(D)$ from the replay memory $\mathcal D$ and trying to
approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by Bellman's equation \eqref{eq:bellman}, namely
$r + \gamma \max_{a'} Q(s', a'|\theta^Q)$, with the output of a parametrized regression model $(s,a,r,s') \mapsto {Q}(s, a|\theta^Q)$, where $Q(.,.|\theta^Q)$ is an approximator for the $Q$-value operator,  parameterized by $\theta^Q$. Computing an optimal policy then corresponds to finding the parameters $\theta^Q$ which minimize the following loss function
\begin{equation}
  \mathcal L(\theta^Q)
  % = \mathbb E_{(s, a, r, s') \sim U(\mathcal D)}[(Q(s, a|\theta^Q) - Q(s',a'|\theta^Q))^2]
  = \mathbb E_{(s, a, r, s') \sim U(\mathcal D)}[(Q(s, a|\theta^Q) - y)^2],
  \label{eq:oracle}
  \end{equation}
where
$ y = r + \gamma \max_{a' \in \mathcal A} Q(s', a'|\theta^Q) = r + \gamma Q(s', \pi(s')|\theta^Q)$ is the prediction target.
% While $y_t$ is also dependent on $\theta^Q$, this is typically ignored.
